---
title: "Jones_Final_Project"
author: "Steve Jones"
date: "May 08, 2017"
output:
  html_document:
    depth: 4
    highlight: tango
    number_sections: no
    theme: united
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---
# Purpose
The input dataset contains information about 710 patient encounters. 50% of the patient within this dataset experience a critical event (Code Blue or Rapid Response) The remaining patients did not have an event.
The primary goal is to predict a critical event using the patients MEWS (Modified Early Warning Score). The secondary goal is to identify actionable trigger points based on MEWS scores.

https://en.wikipedia.org/wiki/Early_warning_score

The Modified Early Warning System (MEWS) is a tool for nurses to help monitor their patients and improve how quickly a patient experiencing a sudden decline receives clinical care. The color-banded patient chart gives a visual cue as to when to calculate a MEWS score.Scoring is based on:
Respiratory rate
Heart rate
Systolic blood pressure
Conscious level
Temperature
Hourly urine output (for previous 2 hours)






# Environment Prep - Install and load packages

##### If package not already installed then install it
```{r, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
if (!require("ggplot2",character.only = TRUE)) (install.packages("ggplot2",dep=TRUE))
if (!require("MASS",character.only = TRUE)) (install.packages("MASS",dep=TRUE))
if (!require("knitr",character.only = TRUE)) (install.packages("knitr",dep=TRUE))
if (!require("xtable",character.only = TRUE)) (install.packages("xtable",dep=TRUE))
if (!require("dplyr",character.only = TRUE)) (install.packages("dplyr",dep=TRUE))
if (!require("psych",character.only = TRUE)) (install.packages("psych",dep=TRUE))
if (!require("stringr",character.only = TRUE)) (install.packages("stringr",dep=TRUE))
if (!require("car",character.only = TRUE)) (install.packages("car",dep=TRUE))
if (!require("faraway",character.only = TRUE)) (install.packages("faraway",dep=TRUE))
if (!require("aod",character.only = TRUE)) (install.packages("aod",dep=TRUE))
if (!require("Rcpp",character.only = TRUE)) (install.packages("Rcpp",dep=TRUE))
if (!require("leaps",character.only = TRUE)) (install.packages("leaps",dep=TRUE))
if (!require("ISLR",character.only = TRUE)) (install.packages("ISLR",dep=TRUE))
if (!require("AUC",character.only = TRUE)) (install.packages("AUC",dep=TRUE))
if (!require("ROCR",character.only = TRUE)) (install.packages("ROCR",dep=TRUE))
if (!require("Amelia",character.only = TRUE)) (install.packages("Amelia",dep=TRUE))
if (!require("popbio",character.only = TRUE)) (install.packages("popbio",dep=TRUE))
if (!require("gdata",character.only = TRUE)) (install.packages("gdata",dep=TRUE))
if (!require("reshape",character.only = TRUE)) (install.packages("reshape",dep=TRUE))
if (!require("rpart",character.only = TRUE)) (install.packages("rpart",dep=TRUE))
if (!require("randomForest",character.only = TRUE)) (install.packages("randomForest",dep=TRUE))
if (!require("ResourceSelection",character.only = TRUE)) (install.packages("ResourceSelection",dep=TRUE))
if (!require("boot",character.only = TRUE)) (install.packages("boot",dep=TRUE))
if (!require("pander",character.only = TRUE)) (install.packages("pander",dep=TRUE))
if (!require("cowplot",character.only = TRUE)) (install.packages("cowplot",dep=TRUE))


##### Load packages
library(plyr)

```


# Data Exploration
  In this section, I load the data and explore it
  
```{r}
dev.off()
setwd("C:/Rockhurst/ADM/Data")
MEWSData <- read.csv("MEW_FinalProject.csv")


str(MEWSData)
names(MEWSData)
summary(MEWSData)
MEWSDATA_EVENTS <- subset(MEWSData, Event==1)

#create table with event volumes by variables
table(MEWSData$Event)
table(MEWSData$Level.Of.Care)
table(MEWSData$Gender)
table(MEWSData$MEWS_4)
table(MEWSData$MEWS_8)
table(MEWSData$MEWS_12)
table(MEWSData$Delta_at_4)
table(MEWSData$T12H_Delta)
table(MEWSData$MEWS4_Tdelta)


library(psych)
pairs.panels(MEWSData)
pairs(MEWSData)


library(lattice)


par(mfrow=c(1,4))

histogram(MEWSData$MEWS_4, type = "count",col="red")
title("MEWS_Pre4")

histogram(MEWSData$MEWS_8, type = "count", col="yellow")
title("MEWS_Pre8")

histogram(MEWSData$MEWS_12, type = "count", col="green")
title("MEWS_Pre12")

histogram(MEWSData$T12H_Delta, type = "count", col="cyan")
title("12HR_Delta")

```


# Data Preparation

 I performed the following transformations on the data either R or directly in Excel:
 
- I randomly selected 80% of the observations as a model training and the remaining 20% was used for testing 

- Create dummy variables
- Data Summary Analysis (R)

```{r}

library(tm)
library(caret)

# Validation approach using 80-20 split
set.seed(123)
#inTrain <- createDataPartition(y=MEWSData$Event, p=0.8, list=FALSE)

#MEWS_train<-MEWSData[inTrain,] 
#MEWS_test<-MEWSData[-inTrain,]  

# Backup
#MEWS_train2 <- MEWS_train
#MEWS_test2 <- MEWS_test

# Data set if I need it
#MEWS_train3 <- MEWS_train
#MEWS_test3 <- MEWS_test
library(class)
set.seed(1234)

#MEWS4_Subset <- MEWSData[,c(1,4,9)]

MEWS4_Subset2 <- MEWSDATA_EVENTS[,c(1,4)]





``` 
# Model Building

 In this section I created my models. The following are the three different approaches that will be used here:

- Model 1: This model will be created by using k-means clustering.
- Model 2: This model will be created by using Naive_Bayes. 
# Model Development - k-MEANS CLUSTERING

```{r}
#finding the optimum no. of clusters
#wssplot <- function(data, nc=15, seed=1234){
 # wss <-(nrow(data)-1)*sum(apply(data,2,var))
#  for(i in 2:nc){
 #   set.seed(seed)
#    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
#  plot(1:nc, wss, type="b", xlab="Number of Clusters",
 #      ylab="Within groups sum of squares", col="red")}

#wssplot(MEWS4_Subset2)

library(class)
set.seed(1234)

MEWS4_Cluster <- kmeans(MEWS4_Subset2,3)
MEWS4_Cluster






```
Using the k-means clustering results - I created another dataset 

```{r}
setwd("C:/Rockhurst/ADM/Data")
MEWS_DS <- read.csv("MEW_FinalProject.csv")

table(MEWS_DS)



MEWS_DS_Working <- MEWS_DS

library(psych)
#### Level Of Care
MEWS_DS_Working<- cbind(MEWS_DS_Working,dummy.code(MEWS_DS_Working$Level.Of.Care))

#Delete original catagorical variable
MEWS_DS_Working$Level.Of.Care <- NULL




##################################################

#### Gender
MEWS_DS_Working<- cbind(MEWS_DS_Working,dummy.code(MEWS_DS_Working$Gender))

#Delete original catagorical variable
MEWS_DS_Working$Gender <- NULL

##########################################

MEWS_DS_Working1 <- MEWS_DS_Working[,c(1:2,7,11:15,17:18)]



MEWS_DS_Working1$Event <- factor(MEWS_DS_Working1$Event)

num.vars <- sapply(MEWS_DS_Working1, is.numeric)
MEWS_DS_Working1[num.vars] <- lapply(MEWS_DS_Working1[num.vars], scale)

```
# Model Development - NAIVE BAYES
 
```{r}
# Validation approach using 80-20 split
set.seed(123)
TRAINING_SET<-data.frame(MEWS_DS_Working1[1:568, ]) 


TEST_SET<-data.frame(MEWS_DS_Working1[569:710, ])  

mews_train_labels<-MEWS_DS_Working1[1:568,1]
mews_test_labels<-MEWS_DS_Working1[569:710,1]


Pred_mews_knn<-knn(train=TRAINING_SET, test=TEST_SET, cl=mews_train_labels, k=27)

library(gmodels)

CrossTable(x=mews_test_labels, y=Pred_mews_knn, prob.chisq=FALSE)

TP = 61
TN = 73
FP = 3
FN = 5

Sensitivity = TP/(TP+FN) #true positive rate; recall; TP/(TP+FN)
Specificity = TN/(TN+FP) #how often is the prediction negative when actual is negative?
Precision = TP/(TP+FP) #how often is prediction positive when actual is positive?
Accuracy = (TP+TN)/(TP+TN+FP+FN)

Value<-round(c(TP,TN,FP,FN,Sensitivity,Specificity,Precision,Accuracy),digits=3)
Measure<-c("True Positive","True Negative","False Positive","False Negative","Sensitivity/Recall=TP/(TN+FP)",
           "Specificity=TN/(TN+TP)","Precision=TP/(TP+FP)","Accuracy=(TP+TN)/total")

table<-as.data.frame(cbind(Measure,Value))

library(knitr)

kable(table)

#############################################################################

library(e1071)
NaivesModel <- naiveBayes(TRAINING_SET, TRAINING_SET$Event)

## Step 4: Evaluating model performance ----
library(caret)
Naives_test_pred <- predict(NaivesModel, TEST_SET)


confusionMatrix(Naives_test_pred, TEST_SET$Event, positive="1")

```






# Final Recommendation

 Cluster analysis indicates possible actionable MEWS point values: 
MEWS of 0-3 - Proposed action continue to monitor as usual
MEWS of 4-5 -  Possible action could be increase monitoring to every hour
MEWS 6+ -  Notify provider   
While the knn model performed at an accuracy of 94% and the Naïve Bayes classification model performed at 74%.  The Naïve bayes performed at 100% in specificity vs 96% specificity of the knn classification model. The knn model also had false negatives vs no false negatives in the the Naïve bayes model. 
The trade-off of specificity over accuracy could result in lives saved and the recommendation is to select the Naïve bayes classification model.

# Incorporate feedback
1. An explanation of the MEW score would be helpful, as I didn't fully understand how it was calculated. I added additional info about the MEWS to the purpose section.


2. Presentation could have been a little more clearer. I'll continue to develop this presentation using the suggestion of performing analysis on All MEWS scores since Nov 2016
I notice a fellow classmate made use of Tableau - I will produce another presentation using Tableau to create the visuals. I'm unable to obtain approval for a new dataset prior to deadline.

